---
title: "Oat-Pea Pea Biomass Estimation, Random Forest"
author: "Peter Hyde"
date: "2025-02-05"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE)
```

<br> <br>

### Library

```{r}
library(readr)
library(readxl)
library(readr)
library(tidyverse)
library(kableExtra)
library(tidymodels)
library(ranger)
library(vip)
library(lme4)

```

<br> <br>

### Import the data sets

```{r}
# T3 data file trial name "Cornell_WinterOatPeaIntercrop_2024_Ithaca"
Cornell_WinterOatPeaIntercrop_2024_Ithaca <- read_csv("data/Cornell_WinterOatPeaIntercrop_2024_Ithaca.csv")

# spectral, subplot averages
spectral_5_12_24_data <- read_csv("output/spectral_5-12-24_data.csv")

# plot meta data, not on T3
Cornell_WinterOatPeaIntercrop_2024_Ithaca_plot_meta <- read_excel("data/Cornell_WinterOatPeaIntercrop_2024_Ithaca_plot_meta.xlsx")


```

<br>

### Create full data set

```{r}
subplot_data <- Cornell_WinterOatPeaIntercrop_2024_Ithaca %>% 
  
  mutate(plot_name = gsub("\\_subplot_.*", "", observationUnitName)) %>% # get plot name without subplot number for joining
  
  filter(observationLevel == "subplot") %>% # keep only subplot level data, most of our data is subplot level
  mutate(subplot_number = str_sub(observationUnitName,-1)) %>% # DANGER this only works if you filter to subplot first
  
  left_join(Cornell_WinterOatPeaIntercrop_2024_Ithaca_plot_meta, join_by(plot_name)) %>% # join T3 data with meta, need inter_crop designation, not found on T3 
  
  left_join(spectral_5_12_24_data, join_by(observationUnitName)) %>% # join with spectral, subplot averages  
  
  filter(germplasmName != "Winter Hayden") # remove winter hayden, it was not planted, plots are pea only

```

<br>

### The full data set

```{r}

colnames(subplot_data)
```
<br>

### Formatting subplot data for model

```{r}
model_data <- subplot_data %>%
  filter(inter_crop %in% c("oat","pea","oat-pea")) %>% # keep only oat accessions, remove barley
  
  select(plot_name,plot_subplot,subplot, "observationUnitName" ,"germplasmName",
         "plotNumber",observationLevel,blockNumber,inter_crop, source,
         `Above ground dry biomass - g|day 136|COMP:0000063`,
         `Weed above ground dry biomass - g|day 136|COMP:0000065`,
         `Pea above ground dry biomass - g|day 136|COMP:0000060`,
         `Pea plant height - cm|day 134|COMP:0000062`,
         `Plant height - cm|day 134|COMP:0000054`,
         contains("mean")) %>% # selecting columns with "mean" in them keeps all the spectral subplot means data
  
  dplyr::rename(oat_biomass = `Above ground dry biomass - g|day 136|COMP:0000063`) %>% 
  dplyr::rename(pea_biomass = `Pea above ground dry biomass - g|day 136|COMP:0000060`) %>% 
  dplyr::rename(oat_height = "Plant height - cm|day 134|COMP:0000054") %>% 
  dplyr::rename(pea_height = "Pea plant height - cm|day 134|COMP:0000062") %>% 
  
  filter(observationUnitName != "Cornell_WinterOatPeaIntercrop_2024_Ithaca-PLOT_49_subplot_3") %>% # remove outlier
  filter(observationUnitName != "Cornell_WinterOatPeaIntercrop_2024_Ithaca-PLOT_264_subplot_1") %>% # remove outlier
  filter(observationUnitName != "Cornell_WinterOatPeaIntercrop_2024_Ithaca-PLOT_110_subplot_1") %>% # remove outlier
  filter(observationUnitName != "Cornell_WinterOatPeaIntercrop_2024_Ithaca-PLOT_197_subplot_1") %>% # remove outlier
  filter(observationUnitName != "Cornell_WinterOatPeaIntercrop_2024_Ithaca-PLOT_197_subplot_2") %>% # remove outlier
  filter(observationUnitName != "Cornell_WinterOatPeaIntercrop_2024_Ithaca-PLOT_197_subplot_3") %>% # remove outlier
  
  as.data.frame() 

model_data  %>% 
  kable(align = "c") %>% 
  kable_classic(full_width=F,position = "left")%>% 
  scroll_box(width = "900px", height = "300px")

```
<br>

### Create the Training Data Set

want the subplots with biomass samples
keep only the columns wanted in model
this model is for oat and oat_pea inter_crop 

```{r}
training_data <- model_data %>% 
  select(oat_biomass,inter_crop,pea_biomass,oat_height,pea_height,contains("mean"))%>% 
  filter(!is.na(oat_biomass)) %>% 
  filter(!is.na(pea_biomass)) %>% 
  filter(inter_crop == "oat-pea") %>% # only use oat-pea plots for this model 
  select(-oat_biomass) # remove oat_biomass from model predicting pea biomass

training_data  %>% 
  kable(align = "c") %>% 
  kable_classic(full_width=F,position = "left")%>% 
  scroll_box(width = "900px", height = "300px")
```

<br>

### Create the Model

```{r}

# set seed
runif(1)*10000  # ran runif once and set seed with result
seed <- 2051

# Split training data 80% / 20%
set.seed(seed)
split <- initial_split(data = training_data, prop = 0.80)

# Training set
train_set <- training(split)

# Testing set
test_set <- testing(split)

# make the recipe,ie the model, it can also have data transformations
recipe <- 
  recipe(pea_biomass ~ ., data = train_set)  ## predicting pea biomass



```


<br>

### Hyperparameter

adjustments are included in rf_spec, tune()

Apart from partitioning the data into training and test sets, 
the training set will be partitioned into 5 data subsets, 
this allows having a variety of partitions to train and measure 
the result of the hyperparameter adjustment process.

```{r}
set.seed(seed)
folds <- vfold_cv(train_set, v = 5)

# Random forest model spec
rf_spec <- rand_forest(
  mtry = tune(), trees = tune(), min_n = tune()) %>% # parameters will be calculated and fine tuned below.
  set_engine("ranger", num.threads = 7, importance = "impurity") %>%  # using ranger package for random forest 
  set_mode("regression") # this is a regression 

# Random forest workflow
rf_Wf <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(rf_spec)

# Space-filling designs grid
rf_grid <- 
  grid_space_filling(
    min_n(), 
    mtry(range = c(4, 10)), # limit the range to what should be good.
    trees(range = c(50, 3000)), 
    size = 100) # why 100?  looks ok, graphs(autoplot(tune_res)) show variability curve  

# use seven cores for this process
doParallel::registerDoParallel(cores = 7) # this will speed up the processing, not entirely needed for this operation.

set.seed(seed)

tune_res <- 
  rf_Wf %>% 
  tune_grid(
    resamples = folds, grid = rf_grid, 
    metrics = metric_set(rmse, mae, rsq)
  )


# finalize model set up
final_rf <- rf_Wf %>% 
  finalize_workflow(
    show_best(x = tune_res, metric = "rmse", n = 1) # optimized to reduce rmse, can pick other metric_set(rmse, mae, rsq) 
  )


final_rs <-
  final_rf %>%
  last_fit(
    split, metrics = metric_set(rmse, mae, rsq)
  )


final_rs %>% 
  collect_metrics()

final_rs %>% 
  extract_fit_parsnip() %>% 
  vip(num_features = 20)  

autoplot(tune_res) 

as.data.frame(final_rs$.predictions) %>% 
ggplot(aes(.pred,pea_biomass))+
  geom_point() +
  geom_smooth(se = FALSE, method = lm)+
  xlab("Predicted")+
  ylab("Observed")+
  scale_color_brewer(palette="Dark2")+
  theme_classic()+
  theme(axis.text.x=element_text(angle = 90, hjust = 1))+
  theme(axis.text=element_text(size=20), axis.title=element_text(size=20,face="bold"))+
  theme(axis.text.x = element_text(face = "bold", color = "black", size = 16))+
  theme(axis.text.y = element_text(face = "bold", color = "black", size = 16))

```
<br>

### Use Model to Predict Biomass

prediction, out of sample 
prediction data set, out of sample, subplot 2
only using data with 4 replicates
```{r}
full_data <- subplot_data %>%
  filter(inter_crop %in% c("oat","pea","oat-pea")) %>% # keep only oat accessions, remove barley
  
  select(plot_name,plot_subplot,subplot, "observationUnitName" ,"germplasmName",
         "plotNumber",observationLevel,blockNumber,inter_crop, source,
         `Above ground dry biomass - g|day 136|COMP:0000063`,
         `Weed above ground dry biomass - g|day 136|COMP:0000065`,
         `Pea above ground dry biomass - g|day 136|COMP:0000060`,
         `Pea plant height - cm|day 134|COMP:0000062`,
         `Plant height - cm|day 134|COMP:0000054`,
         contains("mean")) %>% # selecting columns with "mean" in them keeps all the spectral subplot means data
  
  dplyr::rename(oat_biomass = `Above ground dry biomass - g|day 136|COMP:0000063`) %>% 
  dplyr::rename(pea_biomass = `Pea above ground dry biomass - g|day 136|COMP:0000060`) %>% 
  dplyr::rename(oat_height = "Plant height - cm|day 134|COMP:0000054") %>% 
  dplyr::rename(pea_height = "Pea plant height - cm|day 134|COMP:0000062")

four_reps <- full_data %>% 
  dplyr::group_by(germplasmName) %>% 
  dplyr::count() %>% 
  filter(n==12)# make a list of genotypes with 4 reps, don't use 2 rep genotypes

prediction_data <- full_data %>% 
  filter(subplot == 2) %>% # only make estimates using subplot 2
  filter(germplasmName %in%  c(four_reps$germplasmName)) %>% # remove 2 rep genotypes
  select(observationUnitName,inter_crop,oat_height,pea_height,contains("mean")) %>% 
  filter(inter_crop == "oat-pea") # this model is only for oat-pea plots
  
prediction_data %>% 
  kable(align = "c") %>% 
  kable_classic(full_width=F,position = "left")%>% 
  scroll_box(width = "900px", height = "300px")

final_wf <- final_rs %>%
  extract_workflow()

final_wf # this is the workflow model with the best paramaters 

pred <- predict(final_wf, prediction_data[,-1] ) # removed observationUnitName it isn't part of the model

biomass_predictions <- bind_cols(prediction_data,pred) %>% 
  select(observationUnitName,.pred) %>% 
  dplyr::rename(pred_biomass = .pred)

biomass_predictions %>% 
  kable(align = "c") %>% 
  kable_classic(full_width=F,position = "left")%>% 
  scroll_box(width = "900px", height = "300px")

```

<br>

### Oat Heritability of Pea Biomass Estimates

```{r}
lm_model_factors <- subplot_data %>% 
  select(observationUnitName,blockNumber,germplasmName) %>% 
  mutate(blockNumber = as.character(blockNumber))

df <- biomass_predictions %>% 
  left_join(lm_model_factors)

biomass <- lme4::lmer(`pred_biomass` ~ (1|germplasmName) + (1|blockNumber),
                   data = df)

df_trait <- as.data.frame(VarCorr(biomass))

vg_trait <- df_trait[df_trait$grp == "germplasmName","vcov"]

ve_trait <- df_trait[df_trait$grp == "Residual", "vcov"]

h2_trait <- vg_trait/(vg_trait + ve_trait/(2))

```


```{r}
h2_trait
```

#### output file 
```{r}

# use prediction model to estimate all subplot 2 biomass
# keep genotypes with only 1 rep
# output file for further analysis


output_data <- full_data %>% 
  filter(subplot == 2) %>% # only make estimates using subplot 2
  select(observationUnitName,inter_crop,oat_height,pea_height,contains("mean")) %>% 
  filter(inter_crop == "oat-pea") # this model is only for oat-pea plots


pred_output <- predict(final_wf, output_data[,-1] ) # removed observationUnitName it isn't part of the model


pea_biomass_predictions <- bind_cols(output_data,pred_output) %>% 
  select(observationUnitName,.pred) %>% 
  dplyr::rename(pred_pea_biomass = .pred)


write.csv(pea_biomass_predictions,"output/intercrop_pred_pea_biomass.csv",row.names=FALSE)


```
